Twenty years of dynamic occupancy models: a review of applications and look towards the future

### 2025-01-28 Manuscript.qmd
Updated introduction box

Dynamic occupancy models.

![](images/clipboard-1253177906.png){width="11cm"}

![](images/clipboard-3209412869.png){width="11cm"}
:::

DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.

>>>>

We give a minimal introduction to the canonical dynamic occupancy model as defined in @mackenzie2003. DOMs are hierarchical models which link two sub-models. The first of these describes the ecological process of occupancy, in which sites move between occupied and unoccupied states over time. Occupancy is unobservable, as we cannot always know the true occupancy state. The second sub-model describes the observational process of detection, allowing us to correct for imperfect detection.

#### Data collection

DOMs require hierarchical data to separate the occupancy and detection processes. Occupancy is measured in discrete, time-bound intervals termed *primary occasions*. Within each primary occasion *t*, multiple observations are recorded at the same sites. Importantly, all sites or primary occasions do not require the same number of observations and some number of missing observations are permissible. While secondary occasions are typically repeat observations, other options including same-visit replicates and spatial replicates are also possible (see @guillera-arroita2010).

![](images/clipboard-2772050481.png){width="600" fig-align="center"}

#### Site occupancy sub-model

Occupancy is described during discrete, time-bound intervals termed *primary occasions.* In each primary occasion *t,* each site *i* may exist in either an occupied (z~i,t~ = 1) or unoccupied state (z~i,t~ = 0). In the first primary occasion, occupancy is determined as a Bernoulli trial with initial occupancy probability 𝛙~1~, such that $z_{i,t=1} \sim Bernoulli(\psi_1)$. In subsequent primary occasions, site occupancy is determined by the site's state in the previous primary occasion and probability of colonisation 𝛄 and extinction 𝛆, such that $z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\epsilon)+(1-z_{i,t-1})*\gamma)$. While we present each of 𝛙~1~, 𝛄, and 𝛆 as constants for simplicity, they may be estimated with respect to covariates which are typically included via a logit link.

![](images/clipboard-3349624601.png){width="600" fig-align="center"}

#### Detection sub-model

During any given survey *j,* observers will either observe (y~i,t,j~ = 1) or not observe (y~i,t,j~ = 0) the target species. In each survey, the probability of observing the target species at a site is given as a Bernoulli trial with detection probability *p,* conditional on the site being occupied, such that $y_{i,t,j} \sim Bernoulli(z_{i,t}*p)$. This formulation assumes that no false positive detections exist; that is, observations are never recorded at truly unoccupied sites. As with the occupancy probabilities, *p* may also be estimated using covariates via a logit link.

![](images/clipboard-411045638.png){width="324" fig-align="center"}

#### Fitting DOMs

DOMs are most often fit with maximum likelihood estimation, which identifies the values of 𝛙~1~, 𝛄, 𝛆, and *p* under which the observed data **y** are most probable. Common tools for fitting DOMs include `PRESENCE`, `MARK`, and the r package `unmarked`. DOMs may also be fit in a Bayesian framework, with available packages including `ubms`.
:::

DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our reviews related aspects of model building.
